---
title: "Final lead models"
author: "Erin Franke and Nicholas Di"
date: "4/26/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidycensus)
library(tidyverse)
library(survey)
library(srvyr)
library(sf)
library(ggtext)
library(broom)
#library(spdep)
library(tidymodels)
library(probably)
tidymodels_prefer()
conflicted::conflict_prefer("spautolm", "spatialreg")

load("DataShapefiles/lead_spatial.RData")
river_lakes <- st_read("DataShapefiles/shp_water_lakes_rivers")
river_lakes_big <- river_lakes %>%
  filter(AREA_ACRES >= 500)
roads <- st_read("DataShapefiles/tl_2019_27_prisecroads")
roads_sub <- st_crop(roads,st_bbox(lead_census))
```

# Initial Model: LASS0 logistic model for High Lead 

```{r}
# MODEL RECIPE AND TUNING
set.seed(123)

lasso_log_variables <- lead_census %>%
  mutate(HighLead = as.factor(HighLead), 
         HighLead = relevel(factor(HighLead), ref="0"), 
         testRatio = tested/numChildtestingage) %>% 
  select(HighLead, medageE, medincomeE, NumHouseE, PopE, propFamilyHouseholds, propHomesBuiltPre1950, SSIRecpE, CensusAgeE, tested, MarrCoupleChldU3E, testRatio, GEOID) %>%
  drop_na() %>% st_drop_geometry()

data_cv10 <- vfold_cv(lasso_log_variables, v = 10)

logistic_lasso_spec_tune <- logistic_reg() %>%
    set_engine('glmnet') %>%
    set_args(mixture = 1, penalty = tune()) %>%
    set_mode('classification')

logistic_rec <- recipe(HighLead~medageE + medincomeE + NumHouseE + PopE + propFamilyHouseholds + propHomesBuiltPre1950 + SSIRecpE + CensusAgeE + tested + MarrCoupleChldU3E +testRatio, data = lasso_log_variables) %>%
    step_normalize(all_numeric_predictors()) %>% 
    step_dummy(all_nominal_predictors())

log_lasso_wf <- workflow() %>% 
    add_recipe(logistic_rec) %>%
    add_model(logistic_lasso_spec_tune) 

penalty_grid <- grid_regular(
  penalty(range = c(-10, 10)), 
  levels = 100)

tune_output <- tune_grid( 
  log_lasso_wf, 
  resamples = data_cv10, 
  metrics = metric_set(roc_auc,accuracy),
  control = control_resamples(save_pred = TRUE, event_level = 'second'),
  grid = penalty_grid 
)

autoplot(tune_output) + theme_classic()

# BEST SE PENALTY
best_se_penalty <- select_by_one_std_err(tune_output, metric = 'roc_auc', desc(penalty)) 
best_se_penalty

final_fit_se <- finalize_workflow(log_lasso_wf, best_se_penalty) %>%
    fit(data = lasso_log_variables)

# Variables not driven to 0: medageE, medincomeE, propHomesBuiltPre1950, tested, testRatio
final_fit_se %>% tidy()

# VARIABLE IMPORTANCE
glmnet_output <- final_fit_se %>% extract_fit_engine()
    
bool_predictor_exclude <- glmnet_output$beta==0

var_imp <- sapply(seq_len(nrow(bool_predictor_exclude)), function(row) {
    this_coeff_path <- bool_predictor_exclude[row,]
    ncol(bool_predictor_exclude) - which.min(this_coeff_path) + 1
})

var_imp_data <- tibble(
    var_name = rownames(bool_predictor_exclude),
    var_imp = var_imp
)
var_imp_data %>% arrange(desc(var_imp))
```

```{r}
final_output <- final_fit_se %>% predict(new_data = lasso_log_variables, type='prob') %>% bind_cols(lasso_log_variables)

final_output %>%
    roc_curve(HighLead, .pred_1,event_level = 'second') %>%
    autoplot()

threshold_output <- final_output %>%
    threshold_perf(truth = HighLead, estimate = .pred_0, thresholds = seq(0,1,by=.01)) 

threshold_output %>%
    filter(.metric == 'j_index') %>%
    ggplot(aes(x = .threshold, y = .estimate)) +
    geom_line() +
    labs(y = 'J-index', x = 'threshold') +
    theme_classic()

#Chose a threshold of 0.85 to predict HighLead
threshold_output %>%
    filter(.metric == 'j_index') %>%
    arrange(desc(.estimate))

log_metrics <- metric_set(accuracy,sens,yardstick::spec)

# Model gives an accuracy of 79.69% with a sensitivity of 87.7% and specificity of 77.05%
final_output %>%
    mutate(.pred_class = make_two_class_pred(.pred_0, levels(HighLead), threshold = .85)) %>%
    log_metrics(truth = HighLead, estimate = .pred_class, event_level = 'second')

# Final predictions 
final_lasso_preds <- lead_census %>%
  select(GEOID) %>%
  right_join(final_output, by = "GEOID") %>%
  mutate(.pred_class = make_two_class_pred(.pred_0, levels(HighLead), threshold = .85))
```

Visualize and compare LASSO model predictions and actual result
```{r}
final_lasso_preds %>%
  ggplot() +
  geom_sf(aes(fill = as.factor(.pred_class)))+
  scale_fill_manual(values = c("navy", "red"))+
  theme_classic()+
  geom_sf(data=river_lakes_big, fill="lightblue")+
  labs(title = "Logistic model predictions for lead levels in Twin Cities Metro", fill = "High Lead Predicted")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        legend.title = element_text(family = "mono", size = 9),
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot", 
        legend.text = element_text(family = "mono"))

final_lasso_preds %>%
  ggplot() +
  geom_sf(aes(fill = as.factor(HighLead)))+
  scale_fill_manual(values = c("navy", "red"))+
  theme_classic()+
  geom_sf(data=river_lakes_big, fill="lightblue")+
  labs(title = "Actual high lead census tracts in Twin Cities Metro", fill = "High Lead")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        legend.title = element_text(family = "mono", size = 9),
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot", 
        legend.text = element_text(family = "mono"))
```

# Check LASSO logistic model for spatial correlation using a KNN neighborhood structure 

Create the KNN neighborhood structure and visualize: 
```{r}
library(spdep)
centroids <- st_centroid(st_geometry(final_lasso_preds), of_largest_polygon = TRUE)
knn <- knn2nb(knearneigh(centroids, k = 4))
nb_knn_net <- nb2lines(nb = knn, coords = centroids, as_sf = TRUE)

final_lasso_preds %>%
  ggplot()+
  geom_sf(data = final_lasso_preds, color = "gray", fill="lightblue")+
  geom_sf(data=centroids, lwd = 0.1)+
  geom_sf(data = nb_knn_net)+
  theme_classic()+
  labs(title = "KNN (4) neighborhood structure Twin Cities Metro Area")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot")
```

Use Moran's I test to check for spatial autocorrelation. There is strong evidence for spatial autocorrelation given the small p-value.
```{r}
Wb <- nb2listw(knn, style = "B") #style = 'B' gives binary weights
final_lasso_preds$resid <- as.numeric(as.character(final_lasso_preds$HighLead)) - final_lasso_preds$.pred_1
spdep::moran.test(final_lasso_preds$resid, Wb, alternative = "two.sided", randomisation = TRUE)  # Using randomization test
```

# Create logistic model that accounts for spatial correlation

Get X and Y coordinates for centroids 
```{r}
final_lasso_preds$X <- st_coordinates(centroids)[,1]
final_lasso_preds$Y <- st_coordinates(centroids)[,2]
```

Model One: uses variables from LASSO logistic model with no interactions
```{r}
library(spaMM)
mod1data <- final_lasso_preds %>% 
  mutate(medincomeE = medincomeE/1000,
         propHomesBuiltPre1950 = propHomesBuiltPre1950*100)

# logistic_mod1 <- fitme(HighLead ~ medincomeE + propHomesBuiltPre1950 + testRatio + medageE + Matern(1 | X+Y), fixed = list(nu = 0.5), family = binomial(), data=mod1data,control.HLfit = list(algebra='decorr'))
#logistic_mod1 <- fitme(HighLead ~ medincomeE + propHomesBuiltPre1950 + tested + testRatio + medageE + Matern(1 | X+Y), fixed = list(nu = 0.5), family = binomial(), data=final_lasso_preds,control.HLfit = list(algebra='decorr'))

saveRDS(logistic_mod1, "logistic_mod1_edited.rds")
mod1 <- readRDS("logistic_mod1.rds")
summary(mod1)
mod1_prediction <- final_lasso_preds %>% 
  mutate(logodds = predict(mod1, data = final_lasso_preds),
         odds = exp(logodds),
         probability = odds/(1+odds)) %>% 
  mutate(predHighLead = case_when(probability >= 0.99~1,
                                  probability < 0.99~0)) %>% 
  mutate(Correct = case_when(predHighLead == HighLead ~ 1,
                             predHighLead != HighLead ~ 0)) %>% 
  st_drop_geometry() %>% 
  select(Correct) %>% 
  group_by(Correct) %>% 
  count()

mod1_prediction %>%
  ggplot() +
  geom_sf(aes(fill = as.factor(predHighLead)))+
  scale_fill_manual(values = c("navy", "red"))+
  theme_classic()+
  geom_sf(data=river_lakes_big, fill="lightblue")+
  labs(title = "Fixed Logistic model predictions for lead levels in Twin Cities Metro", fill = "High Lead Predicted")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        legend.title = element_text(family = "mono", size = 9),
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot", 
        legend.text = element_text(family = "mono"))

mod1_prediction %>%
  ggplot() +
  geom_sf(aes(fill = as.factor(HighLead)))+
  scale_fill_manual(values = c("navy", "red"))+
  theme_classic()+
  geom_sf(data=river_lakes_big, fill="lightblue")+
  labs(title = "Actual high lead census tracts in Twin Cities Metro", fill = "High Lead")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        legend.title = element_text(family = "mono", size = 9),
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot", 
        legend.text = element_text(family = "mono"))
```

##TempNotes
Coefficents are same, just as any other model. One unit increase in  medincome. Turning it into something meaningful. prop and tested. the estimate is large enough relative to the standard errors they are far from zero. While others are not too far from zero. will have to do e^17.41 to get odds ratios. multiply prophomespuiltpre1950 by 100 so more interpretable. divide mediam income by 1000. 

correlation parameters: they are rho -> range parameter. how far 2 census tracts are such that they're still corelated. units that are not meaningful. Because i have matern corelation coeffic, does assume isotropic. directional pattern won't capture it. You have log liklihood. can compare it with a logistic regression. should fit better. Accountig through correlation through the random effect. As long as rho is not zero, there is correlation that it is modeling and taking into account. 
##TempNotes

Model Two: includes an interaction between income level and home age
```{r}
mod2data <- final_lasso_preds%>%
  mutate(IncomeCategory = case_when(medincomeE > 0 & medincomeE < 60000 ~ "low",
                                    medincomeE >= 60001 & medincomeE < 80000 ~ "medlow",
                                    medincomeE >= 80001 & medincomeE < 100000 ~ "medhigh",
                                    medincomeE >= 100001 ~ "high"))  %>% 
  mutate(medincomeE = medincomeE/1000,
         propHomesBuiltPre1950 = propHomesBuiltPre1950*100) 

mod2data <- mod2data %>% na.omit()


# logistic_mod2 <- fitme(HighLead ~ IncomeCategory*propHomesBuiltPre1950 + testRatio + medageE + Matern(1 | X+Y), fixed = list(nu = 0.5), family = binomial(), data=mod2data,control.HLfit = list(algebra='decorr'))

saveRDS(logistic_mod2, "logistic_mod2_edited.rds")
mod2 <- readRDS("logistic_mod2_edited.rds")
summary(mod2)

#Getting Error here not sure why
mod2_prediction <- mod2data %>% 
  mutate(logodds = predict(mod2, data = mod2data),
         odds = exp(logodds),
         probability = odds/(1+odds)) %>% 
  mutate(predHighLead = case_when(probability >= 0.7~1,
                                  probability < 0.7~0))

mod2_prediction %>%
  ggplot() +
  geom_sf(aes(fill = as.factor(predHighLead)))+
  scale_fill_manual(values = c("navy", "red"))+
  theme_classic()+
  geom_sf(data=river_lakes_big, fill="lightblue")+
  labs(title = "2nd Fixed Logistic model predictions for lead levels in Twin Cities Metro", fill = "High Lead Predicted")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        legend.title = element_text(family = "mono", size = 9),
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot", 
        legend.text = element_text(family = "mono"))

mod2_prediction %>%
  ggplot() +
  geom_sf(aes(fill = as.factor(HighLead)))+
  scale_fill_manual(values = c("navy", "red"))+
  theme_classic()+
  geom_sf(data=river_lakes_big, fill="lightblue")+
  labs(title = "Actual high lead census tracts in Twin Cities Metro", fill = "High Lead")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        legend.title = element_text(family = "mono", size = 9),
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot", 
        legend.text = element_text(family = "mono"))
```

# Create model for percent (proportion of children with high lead levels)

This model will only use census tracts where the proportion of children with high lead levels is greater than 1%. 

## First, use LASSO to determine predictors for this 
```{r}
lasso_percent_variables <- lead_census %>%
  mutate(HighLead = as.factor(HighLead), 
         HighLead = relevel(factor(HighLead), ref="0"), 
         testRatio = tested/numChildtestingage) %>% 
  select(HighLead, medageE, medincomeE, NumHouseE, PopE, propFamilyHouseholds, propHomesBuiltPre1950, SSIRecpE, CensusAgeE, tested, MarrCoupleChldU3E, testRatio, GEOID, propHomesBuilt1950to1969, propHomesBuilt1970to1989, propHomesBuilt1990to2009, propHomesBuilt2010tonow, percent, propConstruction) %>%
  filter(HighLead == "1") %>%
  drop_na() %>% st_drop_geometry()

set.seed(123)
percent_cv10 <- vfold_cv(lasso_percent_variables, v = 10)

lm_lasso_spec_tune <- 
  linear_reg() %>%
  set_args(mixture = 1, penalty = tune()) %>% 
  set_engine(engine = 'glmnet') %>%
  set_mode('regression') 

percent_rec <- recipe(percent ~., data = lasso_percent_variables) %>%
  update_role(GEOID, new_role = "ID") %>% # we don't want to use ID as predictor
  step_novel(all_nominal_predictors()) %>% 
  step_dummy(all_nominal_predictors()) %>% 
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())  # important step for LASSO

lasso_wf_percent <- workflow() %>% 
  add_recipe(percent_rec) %>%
  add_model(lm_lasso_spec_tune) 

penalty_grid <- grid_regular(
  penalty(range = c(-5, 3)), #log10 transformed 10^-5 to 10^3
  levels = 50)

tune_res <- tune_grid( # new function for tuning hyperparameters
  lasso_wf_percent, # workflow
  resamples = percent_cv10, # folds
  metrics = metric_set(rmse),
  grid = penalty_grid # penalty grid
)

autoplot(tune_res)

collect_metrics(tune_res) %>%
  filter(.metric == 'rmse') %>%
  select(penalty, rmse = mean) 

best_penalty <- select_best(tune_res, metric = 'rmse', desc(penalty)) 
percent_final_wk <- finalize_workflow(lasso_wf_percent, best_penalty) # incorporates penalty value to workflow
percent_final_fit <- fit(percent_final_wk, data = lasso_percent_variables)
tidy(percent_final_fit)

percent_final_fit %>% predict(new_data = lasso_percent_variables)

# we can predict the percent of high lead children on average within 1.5% 
tune_res %>% collect_metrics() %>% filter(penalty == (best_penalty %>% pull(penalty)))

lasso_mod_out <- percent_final_fit %>%
    predict(new_data = lasso_percent_variables) %>%
    bind_cols(lasso_percent_variables) %>%
    mutate(resid = percent - .pred)

lasso_mod_out %>% 
  ggplot(aes(x = .pred, y = resid)) + 
  geom_point() +
  geom_smooth(se = FALSE) + 
  geom_hline(yintercept = 0) + 
  theme_classic()
```

## Moran's I test for percent 

Look for evidence of spatial correlation in residuals
```{r}
high_metro_centroids <- st_centroid(st_geometry(lasso_results), of_largest_polygon = TRUE)
knn <- knn2nb(knearneigh(high_metro_centroids, k = 4))
nb_knn_net <- nb2lines(nb = knn, coords = high_metro_centroids, as_sf = TRUE)

lasso_results <- lead_census %>%
  filter(HighLead == 1) %>%
  select(GEOID) %>%
  right_join(lasso_mod_out, by = "GEOID")

Wb <- nb2listw(knn, style = "B") #style = 'B' gives binary weights
spdep::moran.test(lasso_results$resid, Wb, alternative = "two.sided", randomisation = TRUE)  # Using randomization test
```

Given spatial correlation in residuals, create a KNN neighborhood structure
```{r}
lasso_results %>%
  ggplot()+
  geom_sf(data = lasso_results, color = "gray", fill="lightblue")+
  geom_sf(data=high_metro_centroids, lwd = 0.1)+
  geom_sf(data = nb_knn_net)+
  theme_classic()+
  labs(title = "KNN (4) neighborhood structure for high lead centroids")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot")
```

## SAR and CAR models using variables from LASSO

Fit the SAR model for percent using the variables from the LASS0. It appears that the residuals are independent as the Moran's I test gives a p-value of 0.8246.  
```{r}
library(spatialreg)

Ww <- nb2listw(knn, style = "W")
Wb <- nb2listw(knn, style = "B")

mod_sar <- spautolm(formula = percent~ PopE + propHomesBuiltPre1950 + SSIRecpE + propHomesBuilt1950to1969, data = lasso_results, listw = Ww, family = "SAR")
summary(mod_sar)
BIC(mod_sar)

lasso_results$mod_sar_knn <- resid(mod_sar)
spdep::moran.test(lasso_results$mod_sar_knn, Wb, alternative = "two.sided", randomisation = TRUE)
```

Fit the CAR model for percent using the variables from the LASS0. The p-value for the Moran's I test for the car model is 0.2527. Additionally, the BIC is more negative for the SAR model (-562.28 for SAR and -554.87 for CAR) and so it seems that the SAR model is the better choice. 
```{r}
mod_car <- spautolm(formula = percent~ PopE + propHomesBuiltPre1950 + SSIRecpE + propHomesBuilt1950to1969, data = lasso_results, listw = Ww, family = "CAR")
summary(mod_car)
BIC(mod_car)

lasso_results$mod_car_knn <- resid(mod_car)
spdep::moran.test(lasso_results$mod_car_knn, Wb, alternative = "two.sided", randomisation = TRUE)
```

SAR Model 1 Evaluation. Create maps of predicted and actual percent and a map of the residuals. 
```{r}
lasso_results$sar_pred <- fitted(mod_sar)

lasso_results %>%
  select(sar_pred) %>%
  st_join(lead_census,.) %>%
    ggplot() + 
  geom_sf(size=.05,fill='white') + 
  geom_sf(data = lasso_results,aes(fill = sar_pred),size=.5) + 
  labs(title = "Percent of Children with High Lead Blood Levels: SAR Predictions", fill = "") + 
  scale_fill_gradient(low = "white", high = "darkred", limits = c(0,0.12)) + theme_classic()+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot")

lasso_results %>%
    ggplot() + 
  geom_sf(aes(fill = percent)) + 
  labs(title = "Percent of Children with High Lead Blood Levels: Actual", fill = "") + 
  scale_fill_gradient(low = "white", high = "darkred") + 
  theme_classic()+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot")

# Model Residuals 
lasso_results %>%
    ggplot() + 
  geom_sf(aes(fill = mod_sar_knn)) + 
  labs(title = "Residuals from SAR Model", fill = "Residuals") + 
  scale_fill_gradient2(mid = "white", high = "darkred", low = "navy") +
  theme_classic()+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot")
```

Find the RMSE of the SAR model: 0.01468
```{r}
sqrt(mean(lasso_results$mod_car_knn^2))
```

## Potentially another SAR model with different variables here