---
title: "Lead Modeling"
author: "Erin Franke"
date: "4/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidycensus)
library(tidyverse)
library(survey)
library(srvyr)
library(sf)
library(ggtext)
library(broom)
#library(spdep)
library(tidymodels)
library(probably)
tidymodels_prefer()
conflicted::conflict_prefer("spautolm", "spatialreg")

load("DataShapefiles/lead_spatial.RData")
river_lakes <- st_read("DataShapefiles/shp_water_lakes_rivers")
roads <- st_read("DataShapefiles/tl_2019_27_prisecroads")
roads_sub <- st_crop(roads,st_bbox(lead_census))
```

# LASS0 logistic model for High Lead 

```{r}
# MODEL RECIPE AND TUNING
set.seed(123)

lasso_log_variables <- lead_census %>%
  mutate(HighLead = as.factor(HighLead), 
         HighLead = relevel(factor(HighLead), ref="0")) %>% 
  select(HighLead, medageE, medincomeE, NumHouseE, PopE, propFamilyHouseholds, propHomesBuiltPre1950, SSIRecpE, CensusAgeE, tested, MarrCoupleChldU3E) %>%
  drop_na() %>% st_drop_geometry()

data_cv10 <- vfold_cv(lasso_log_variables, v = 10)

logistic_lasso_spec_tune <- logistic_reg() %>%
    set_engine('glmnet') %>%
    set_args(mixture = 1, penalty = tune()) %>%
    set_mode('classification')

logistic_rec <- recipe(HighLead~., data = lasso_log_variables) %>%
    step_normalize(all_numeric_predictors()) %>% 
    step_dummy(all_nominal_predictors())

log_lasso_wf <- workflow() %>% 
    add_recipe(logistic_rec) %>%
    add_model(logistic_lasso_spec_tune) 

penalty_grid <- grid_regular(
  penalty(range = c(-10, 10)), 
  levels = 100)

tune_output <- tune_grid( 
  log_lasso_wf, 
  resamples = data_cv10, 
  metrics = metric_set(roc_auc,accuracy),
  control = control_resamples(save_pred = TRUE, event_level = 'second'),
  grid = penalty_grid 
)

autoplot(tune_output) + theme_classic()

# BEST SE PENALTY
best_se_penalty <- select_by_one_std_err(tune_output, metric = 'roc_auc', desc(penalty)) 
best_se_penalty

final_fit_se <- finalize_workflow(log_lasso_wf, best_se_penalty) %>%
    fit(data = lasso_log_variables)

final_fit_se %>% tidy()

# VARIABLE IMPORTANCE
glmnet_output <- final_fit_se %>% extract_fit_engine()
    
bool_predictor_exclude <- glmnet_output$beta==0

var_imp <- sapply(seq_len(nrow(bool_predictor_exclude)), function(row) {
    this_coeff_path <- bool_predictor_exclude[row,]
    ncol(bool_predictor_exclude) - which.min(this_coeff_path) + 1
})

var_imp_data <- tibble(
    var_name = rownames(bool_predictor_exclude),
    var_imp = var_imp
)
var_imp_data %>% arrange(desc(var_imp))
```

```{r}
final_output <- final_fit_se %>% predict(new_data = lasso_log_variables, type='prob') %>% bind_cols(lasso_log_variables)

final_output %>%
    roc_curve(HighLead, .pred_1,event_level = 'second') %>%
    autoplot()

threshold_output <- final_output %>%
    threshold_perf(truth = HighLead, estimate = .pred_0, thresholds = seq(0,1,by=.01)) 

# J-index v. threshold for not_spam
threshold_output %>%
    filter(.metric == 'j_index') %>%
    ggplot(aes(x = .threshold, y = .estimate)) +
    geom_line() +
    labs(y = 'J-index', x = 'threshold') +
    theme_classic()

threshold_output %>%
    filter(.metric == 'j_index') %>%
    arrange(desc(.estimate))

log_metrics <- metric_set(accuracy,sens,yardstick::spec)

final_output %>%
    mutate(.pred_class = make_two_class_pred(.pred_0, levels(HighLead), threshold = .84)) %>%
    log_metrics(truth = HighLead, estimate = .pred_class, event_level = 'second')
```

# Normal logistic regression with variables LASSO deemed important 

```{r}
lead_census_clean <- lead_census %>%
  filter(!is.na(medincomeE), !is.na(medageE), !is.na(propHomesBuiltPre1950), !is.na(tested))
logistic_mod1 <- glm(HighLead~ medincomeE + medageE + propHomesBuiltPre1950 + tested, data = lead_census_clean, family = binomial())
logistic_mod1 %>%
  coef() %>%
  exp()
tidy(logistic_mod1)

lead_census_clean$temp <- predict(logistic_mod1, type = "response")
lead_census_clean$lm_resid <- resid(logistic_mod1)
lead_census_clean <- lead_census_clean %>%
  mutate(hard_pred = case_when(temp >= 0.16 ~ 1, 
                               TRUE ~ 0))

lead_census_clean <- lead_census_clean %>% 
  mutate(WeAreRight = case_when(as.numeric(HighLead) == hard_pred ~ 1,
                                TRUE ~ 0))

lead_census_clean %>% st_drop_geometry() %>% 
  group_by(WeAreRight) %>% count()

# Model predictions
lead_census_clean %>% 
  ggplot() +
  geom_sf(aes(fill = as.factor(hard_pred)))+
  #scale_fill_gradient2(mid = "white", high = "red", low = "blue",labels = scales::comma) +
  scale_fill_manual(values = c("navy", "red"))+
  theme_classic()+
  #geom_sf(data=areawater, fill="navy")+
  labs(title = "Logistic model predictions for lead levels in Twin Cities Metro", fill = "High Lead Predicted")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        legend.title = element_text(family = "mono", size = 9),
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot", 
        legend.text = element_text(family = "mono"))





lead_census_clean %>% 
  ggplot() +
  geom_sf(aes(fill = as.factor(WeAreRight)),size =.1)+
  #scale_fill_gradient2(mid = "white", high = "red", low = "blue",labels = scales::comma) +
  scale_fill_manual(values = c( "red","grey"))+
  theme_classic()+
  #geom_sf(data=areawater, fill="navy")+
  labs(title = "Correct logistic model predictions for lead levels in Twin Cities Metro", fill = "Correctly Predicted High Lead")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        legend.title = element_text(family = "mono", size = 9),
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot", 
        legend.text = element_text(family = "mono"))


lead_census_clean %>% 
  ggplot() +
  geom_sf(aes(fill = lm_resid),size =.1)+
  scale_fill_gradient2(mid = "white", high = "red", low = "blue",labels = scales::comma) +
  #scale_fill_manual(values = c( "red","grey"))+
  theme_classic()+
  #geom_sf(data=areawater, fill="navy")+
  labs(title = "Prediction Errors for lead levels in Twin Cities Metro", fill = "Residuals")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        legend.title = element_text(family = "mono", size = 9),
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot", 
        legend.text = element_text(family = "mono"))
```

# Now adjusting for spatial correlation

Create a rook neighborhood structure and get the centroids of each census tract.

```{r}

rook <- poly2nb(lead_census_clean, queen = FALSE)
metro_centroids <- st_centroid(st_geometry(lead_census_clean), of_largest_polygon = TRUE)
nb_rook_net <- nb2lines(nb = rook, coords = metro_centroids, as_sf = TRUE)
lead_census_clean %>%
  ggplot()+
  geom_sf(data = lead_census_clean, color = "gray", fill="lightblue")+
  geom_sf(data=metro_centroids, lwd = 0.1)+
  geom_sf(data = nb_rook_net,size=.2)+
  theme_classic()+
  #geom_sf(data=areawater, fill="navy")+
  labs(title = "Rook neighborhood structure for Twin Cities Metro Area")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot")
```

Strong evidence of spatial autocorrelation in logistic model given the very low p-value. 

```{r}
Wb <- nb2listw(rook, style = "B") #style = 'B' gives binary weights
spdep::moran.test(lead_census_clean$lm_resid, Wb, alternative = "two.sided", randomisation = TRUE)  # Using randomization test
```

Thinking about logistic CAR and SAR models to account for spatial autocorrelation - need to figure out how to do logistic model (it runs but not sure it is thinking of 0 and 1 as factor)

```{r}
library(spatialreg)
# Convert Neighborhood Information to List (with weighting so that rows sum to 1)
Ww <- nb2listw(rook, style = "W")
# Fit SAR Model - this run but need to figure out it is the correct way for a logistic model
mod_sar <- spatialreg::spautolm(formula = HighLead ~  medincomeE + medageE + propHomesBuiltPre1950 + tested, data = lead_census_clean, listw = Ww, family = "SAR")
summary(mod_sar)
BIC(mod_sar)


lead_census_clean %>%
  mutate(pred = fitted(mod_sar)) %>%
  ggplot(aes(x = pred,fill = factor(HighLead))) + 
  geom_density(alpha=.2)

lead_census_clean %>%
  mutate(pred = fitted(mod_sar), resid = HighLead - pred) %>%
  ggplot(aes(x = tested,y = resid)) + 
  geom_point(alpha=.2)

lead_census_clean %>%
  mutate(pred = fitted(mod_sar), resid = HighLead - pred) %>%
  ggplot(aes(x = propHomesBuiltPre1950,y = pred, color = factor(HighLead))) + 
  geom_point(alpha=.2) + facet_wrap(~cut(medincomeE,3))
#Consider an interaction with income and old homes. You may want to categorize income.


```

# Model for lead levels of census tracts that have high levels 

```{r}
high_lead_clean <- lead_census_clean %>%
  filter(HighLead == 1) 
highlead_mod1 <- lm(percent~  medincomeE + medageE + propHomesBuiltPre1950 + tested, data = high_lead_clean)
high_lead_clean$lm_resid <- resid(highlead_mod1)

high_lead_clean %>% 
  ggplot() +
  geom_sf(aes(fill = lm_resid)) +
  scale_fill_gradient2(mid = "white", high = "red", low = "blue",labels = scales::comma) +
  theme_classic()+
  #geom_sf(data=areawater, fill="navy")+
  labs(title = "OLS model residuals for high lead levels in Twin Cities Metro", fill = "Residuals")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        legend.title = element_text(family = "mono", size = 9),
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot", 
        legend.text = element_text(family = "mono"))
```

Create KNN neighborhood structure (Rook some will have no neighbors)
```{r}
high_metro_centroids <- st_centroid(st_geometry(high_lead_clean), of_largest_polygon = TRUE)
knn <- knn2nb(knearneigh(high_metro_centroids, k = 4))
nb_knn_net <- nb2lines(nb = knn, coords = high_metro_centroids, as_sf = TRUE)

high_lead_clean %>%
  ggplot()+
  geom_sf(data = high_lead_clean, color = "gray", fill="lightblue")+
  geom_sf(data=high_metro_centroids, lwd = 0.1)+
  geom_sf(data = nb_knn_net)+
  theme_classic()+
  #geom_sf(data=areawater, fill="navy")+
  labs(title = "KNN (4) neighborhood structure for high lead centroids")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot")
```

OLS model shows evidence of spatial correlation. 
```{r}
Wb <- nb2listw(knn, style = "B") #style = 'B' gives binary weights
spdep::moran.test(high_lead_clean$lm_resid, Wb, alternative = "two.sided", randomisation = TRUE)  # Using randomization test
```

```{r}
library(spatialreg)
# Convert Neighborhood Information to List (with weighting so that rows sum to 1)
Ww <- nb2listw(knn, style = "W")
Wb <- nb2listw(knn, style = "B")
# Fit SAR Model - this run but need to figure out it is the correct way for a logistic model
mod_sar <- spautolm(formula = percent~ propHomesBuiltPre1950 + medincomeE + PopE, data = high_lead_clean, listw = Ww, family = "SAR")
summary(mod_sar)
BIC(mod_sar)

high_lead_clean$mod_sar_knn <- resid(mod_sar)
spdep::moran.test(high_lead_clean$mod_sar_knn, Wb, alternative = "two.sided", randomisation = TRUE)

# fit the car model
mod_car <- spautolm(formula = percent~ medincomeE + propHomesBuiltPre1950 + PopE, data = high_lead_clean, listw = Ww, family = "CAR")
summary(mod_car)
BIC(mod_car)

high_lead_clean$mod_car_knn <- resid(mod_car)
spdep::moran.test(high_lead_clean$mod_car_knn, Wb, alternative = "two.sided", randomisation = TRUE)
```



##Nick Trying KNN with logistic model for HighLead

Creating KNN Network 

```{r}
lead_census_clean$X <- st_coordinates(metro_centroids)[,1]
lead_census_clean$Y <- st_coordinates(metro_centroids)[,2]
#Will have X and Y cord for fit me
```

Modeling SAR Under KNN 
```{r}
#Add X and Y Cord to Census Tract 

Ww <- nb2listw(knn, style = "W")
Wb <- nb2listw(knn, style = "B")

# Fit SAR Model - this run but need to figure out it is the correct way for a logistic model
mod_sar_knn <- spautolm(formula = HighLead~ medincomeE + propHomesBuiltPre1950 + PopE, data = lead_census_clean, listw = Ww, family = "SAR")
summary(mod_sar_knn)
BIC(mod_sar_knn)

library(spaMM)

logistic_attempt <- fitme(HighLead ~ medincomeE + propHomesBuiltPre1950 + PopE + Matern(1 | X+Y), fixed = list(nu = 0.5), family = binomial(), data=lead_census_clean,control.HLfit = list(algebra='decorr'))
saveRDS(logistic_attempt, "Logistic_model.rds")
my_model <- readRDS("Logistic_model.rds")

summary(my_model)

 # Fit CAR Model - this run but need to figure out it is the correct way for a logistic model
mod_car_knn <- spautolm(formula = HighLead~ medincomeE + propHomesBuiltPre1950 + PopE, data = lead_census_clean, listw = Ww, family = "CAR")
summary(mod_car_knn)
BIC(mod_car_knn)
```

```{r}
high_lead_clean$sar_resid_knn <- resid(mod_sar_knn)
spdep::moran.test(high_lead_clean$sar_resid_knn, Wb, alternative = "two.sided", randomisation = TRUE)

high_lead_clean$car_resid_knn <- resid(mod_car_knn)
spdep::moran.test(high_lead_clean$car_resid_knn, Wb, alternative = "two.sided", randomisation = TRUE)
```







