---
title: "Lead Modeling"
author: "Erin Franke"
date: "4/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidycensus)
library(tidyverse)
library(survey)
library(srvyr)
library(sf)
library(ggtext)
library(broom)
library(spdep)

load("DataShapefiles/lead_spatial.RData")
river_lakes <- st_read("DataShapefiles/shp_water_lakes_rivers")
roads <- st_read("DataShapefiles/tl_2019_27_prisecroads")
roads_sub <- st_crop(roads,st_bbox(lead_census))
```

# Logistic model for High Lead 

Try out a logistic model for `HighLead`:
```{r}
lead_census_clean <- lead_census %>%
  filter(!is.na(medincomeE), !is.na(propHomesBuiltPre1950), !is.na(PopE))
logistic_mod1 <- glm(HighLead~ medincomeE + propHomesBuiltPre1950 + PopE, data = lead_census_clean, family = binomial())
logistic_mod1 %>%
  coef() %>%
  exp()
tidy(logistic_mod1)
lead_census_clean$lm_resid <- resid(logistic_mod1)

lead_census_clean %>% 
  ggplot() +
  geom_sf(aes(fill = lm_resid)) +
  scale_fill_gradient2(mid = "white", high = "red", low = "blue",labels = scales::comma) +
  theme_classic()+
  #geom_sf(data=areawater, fill="navy")+
  labs(title = "Logistic model residuals for lead levels in Twin Cities Metro", fill = "Residuals")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        legend.title = element_text(family = "mono", size = 9),
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot", 
        legend.text = element_text(family = "mono"))
```

Rook neighborhood structure
```{r}
rook <- poly2nb(lead_census_clean, queen = FALSE)
metro_centroids <- st_centroid(st_geometry(lead_census_clean), of_largest_polygon = TRUE)
nb_rook_net <- nb2lines(nb = rook, coords = metro_centroids, as_sf = TRUE)
lead_census_clean %>%
  ggplot()+
  geom_sf(data = lead_census_clean, color = "gray", fill="lightblue")+
  geom_sf(data=metro_centroids, lwd = 0.1)+
  geom_sf(data = nb_rook_net)+
  theme_classic()+
  #geom_sf(data=areawater, fill="navy")+
  labs(title = "Rook neighborhood structure for Twin Cities Metro Area")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot")
```

Strong evidence of spatial autocorrelation in logistic model:
```{r}
Wb <- nb2listw(rook, style = "B") #style = 'B' gives binary weights
spdep::moran.test(lead_census_clean$lm_resid, Wb, alternative = "two.sided", randomisation = TRUE)  # Using randomization test
```

Thinking about logistic CAR and SAR models to account for spatial autocorrelation - need to figure out how to do logistic model (it runs but not sure it is thinking of 0 and 1 as factor)
```{r}
library(spatialreg)
# Convert Neighborhood Information to List (with weighting so that rows sum to 1)
Ww <- nb2listw(rook, style = "W")
# Fit SAR Model - this run but need to figure out it is the correct way for a logistic model
mod_sar <- spautolm(formula = HighLead~ medincomeE + propHomesBuiltPre1950 + PopE, data = lead_census_clean, listw = Ww, family = "SAR")
summary(mod_sar)
BIC(mod_sar)

# fit the car model
mod_car <- spautolm(formula = HighLead~ medincomeE + propHomesBuiltPre1950 + PopE, data = lead_census_clean, listw = Ww, family = "CAR")
BIC(mod_car)
```

# Model for lead levels of census tracts that have high levels 

```{r}
high_lead_clean <- lead_census %>%
  filter(!is.na(medincomeE), !is.na(propHomesBuiltPre1950), !is.na(PopE), HighLead == 1)
highlead_mod1 <- lm(percent~ medincomeE + propHomesBuiltPre1950 + PopE, data = high_lead_clean)
high_lead_clean$lm_resid <- resid(highlead_mod1)

high_lead_clean %>% 
  ggplot() +
  geom_sf(aes(fill = lm_resid)) +
  scale_fill_gradient2(mid = "white", high = "red", low = "blue",labels = scales::comma) +
  theme_classic()+
  #geom_sf(data=areawater, fill="navy")+
  labs(title = "OLS model residuals for high lead levels in Twin Cities Metro", fill = "Residuals")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        legend.title = element_text(family = "mono", size = 9),
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot", 
        legend.text = element_text(family = "mono"))
```

Create rook neighborhood structure 
```{r}
high_metro_centroids <- st_centroid(st_geometry(high_lead_clean), of_largest_polygon = TRUE)
knn <- knn2nb(knearneigh(high_metro_centroids, k = 4))
nb_knn_net <- nb2lines(nb = knn, coords = high_metro_centroids, as_sf = TRUE)

high_lead_clean %>%
  ggplot()+
  geom_sf(data = high_lead_clean, color = "gray", fill="lightblue")+
  geom_sf(data=high_metro_centroids, lwd = 0.1)+
  geom_sf(data = nb_knn_net)+
  theme_classic()+
  #geom_sf(data=areawater, fill="navy")+
  labs(title = "Rook neighborhood structure for high lead centroids")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot")
```

OLS model shows evidence of spatial correlation. 
```{r}
Wb <- nb2listw(knn, style = "B") #style = 'B' gives binary weights
spdep::moran.test(high_lead_clean$lm_resid, Wb, alternative = "two.sided", randomisation = TRUE)  # Using randomization test
```

```{r}
library(spatialreg)
# Convert Neighborhood Information to List (with weighting so that rows sum to 1)
Ww <- nb2listw(knn, style = "W")
# Fit SAR Model - this run but need to figure out it is the correct way for a logistic model
mod_sar <- spautolm(formula = percent~ propHomesBuiltPre1950 + medincomeE + PopE, data = high_lead_clean, listw = Ww, family = "SAR")
summary(mod_sar)
BIC(mod_sar)

high_lead_clean$mod_sar_knn <- resid(mod_sar)
spdep::moran.test(high_lead_clean$mod_sar_knn, Wb, alternative = "two.sided", randomisation = TRUE)

# fit the car model
mod_car <- spautolm(formula = HighLead~ medincomeE + propHomesBuiltPre1950 + PopE, data = lead_census_clean, listw = Ww, family = "CAR")
BIC(mod_car)
```



##Nick Trying KNN

Creating KNN Network 

```{r}
knn <- knn2nb(knearneigh(metro_centroids, k = 4))
nb_knn_net <- nb2lines(nb = knn, coords = metro_centroids, as_sf = TRUE)

lead_census_clean$X <- st_coordinates(metro_centroids)[,1]
lead_census_clean$Y <- st_coordinates(metro_centroids)[,2]
#Will have X and Y cord for fit me


lead_census_clean %>%
  ggplot()+
  geom_sf(data = lead_census_clean, color = "gray", fill="lightblue")+
  geom_sf(data=metro_centroids, lwd = 0.1)+
  geom_sf(data = nb_knn_net)+
  theme_classic()+
  #geom_sf(data=areawater, fill="navy")+
  labs(title = "KNN neighborhood structure for Twin Cities Metro Area")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(), 
        plot.title = element_text(family = "mono", size =11), 
        plot.title.position = "plot")
```

```{r}
Wb <- nb2listw(knn, style = "B") #style = 'B' gives binary weights
spdep::moran.test(lead_census_clean$lm_resid, Wb, alternative = "two.sided", randomisation = TRUE)

#Error Here Note Sure
```

Modeling SAR Under KNN 
```{r}
#Add X and Y Cord to Census Tract 

Ww <- nb2listw(knn, style = "W")
Wb <- nb2listw(knn, style = "B")

# Fit SAR Model - this run but need to figure out it is the correct way for a logistic model
mod_sar_knn <- spautolm(formula = HighLead~ medincomeE + propHomesBuiltPre1950 + PopE, data = lead_census_clean, listw = Ww, family = "SAR")
summary(mod_sar_knn)
BIC(mod_sar_knn)

library(spaMM)

logistic_attempt <- fitme(HighLead ~ medincomeE + propHomesBuiltPre1950 + PopE + Matern(1 | X+Y), fixed = list(nu = 0.5), family = binomial(), data=lead_census_clean,control.HLfit = list(algebra='decorr'))
saveRDS(logistic_attempt, "Logistic_model.rds")
my_model <- readRDS("Logistic_model.rds")

summary(my_model)

 # Fit CAR Model - this run but need to figure out it is the correct way for a logistic model
mod_car_knn <- spautolm(formula = HighLead~ medincomeE + propHomesBuiltPre1950 + PopE, data = lead_census_clean, listw = Ww, family = "CAR")
summary(mod_car_knn)
BIC(mod_car_knn)
```

```{r}
high_lead_clean$sar_resid_knn <- resid(mod_sar_knn)
spdep::moran.test(high_lead_clean$sar_resid_knn, Wb, alternative = "two.sided", randomisation = TRUE)

high_lead_clean$car_resid_knn <- resid(mod_car_knn)
spdep::moran.test(high_lead_clean$car_resid_knn, Wb, alternative = "two.sided", randomisation = TRUE)
```







