---
title: "A Spatial Analysis of Elevated Blood Lead Levels in the Twin Cities Metropolitan Region"
author: "Erin Franke and Nicholas Di"
date: "May 6, 2022"
output: github_document
bibliography: Library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

```{r, message=FALSE, warning=FALSE}
library(tidycensus)
library(tidyverse)
library(survey)
library(srvyr)
library(sf)
library(ggtext)

load("DataShapefiles/lead_spatial.RData")
river_lakes <- st_read("DataShapefiles/shp_water_lakes_rivers")
roads <- st_read("DataShapefiles/tl_2019_27_prisecroads")
roads_sub <- st_crop(roads,st_bbox(lead_census))
```

# Introduction 

When raising a child, parents go through lots of stress to keep their children safe and healthy. From using carseats to getting children vaccinated to working on speech and mobility development and beyond, parents have a lot on their plate. But one thing that may be overlooked in providing safe and healthy environment for a child is lead. Lead in paint, soil, air, or water is invisible to the naked eye and has no smell [@cdcPrevention]. However, children can be exposed to lead in a variety of manners, including swallowing house dust or soil contaminated by lead paint or drinking water delivered through lead-based pipes, faucets, and plumbing fixtures. Exposure to this hidden element can seriously harm a child's health, including damage to the child's brain and nervous system, slowed growth and development, as well as learning, hearing, speech, and behavior problems [@cdcPrevention]. If exposed to especially high levels of lead, children can face a brain condition known as encephalopathy, severe neurological damage, coma, and even death [@leadoverview]. Thus, without a question it is crucial to keep lead exposure to a minimum when raising a child. 

In this analysis, we analyzed elevated blood lead levels in the **7-county Twin Cities metropolitan area** using public data provided by the **Minnesota Department of Health** over the period of 2015-2019 [@mdhdata]. To protect the privacy of individuals, the smallest granularity we were able to obtain this data was on the census tract level, meaning for each of the 691 census tracts in the Twin Cities metropolitan area we obtained information on how many children were tested and how many of those tests resulted in elevated blood lead levels. To have **elevated blood lead levels (EBLL)** means that a child has a confirmed result **at or above 5 micrograms of lead per deciliter of blood (mcg/dL)** [@leadoverview]. Children under 6 years of age are tested. The Minnesota Department of Health idenifies children living in the Minneapolis and Saint Paul city limits as children at a higher risk for lead exposure and recommends these children to receive blood lead testing at 1 and 2 years of age. This recommendation is warranted given that in 2019, between 1-2% of children in Minneapolis or St. Paul had an EBLL, which is double the statewide average and higher than any other region of Minnesota [@leadoverview]. Interestingly, the MDH has found children living in the Metro area but not living in the cities of Minneapolis or St. Paul are at a lower risk of lead exposure than the Greater Minnesota (non-Metro) are. Only about 0.3% of these children have high EBLL levels whereas about 0.8% of children living in MN outside the metro area have high EBLL levels. As a result, to best explore this contrast between Minneapolis-Saint Paul and the suburban region, this project will solely focus on EBLL data from the 7 county Twin Cities metro area. This region is shown in navy on the road map of Minnesota below. 

```{r}
lead_census %>%
  ggplot()+
  geom_sf(fill = "navy", lwd = 0.2, color = "navy")+
  geom_sf(data = roads, fill = "lightblue",color = "lightblue", lwd=0.2)+
  theme_classic()+
  labs(title = "Twin Cities Metropolitan region on MN road map")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        legend.position = "none", 
        plot.title.position = "plot", 
        plot.title = element_markdown(family = "mono"), 
        plot.subtitle = element_markdown(family = "mono", size = 8))+
  scale_fill_manual(values = c("navy", "red"))
```

# Research Goal

Keeping the health consequences of lead exposure to children in the front of our minds, our research focuses on investigating what is correlated with a census tract having a noticeably high proportion of children testing with elevated blood lead levels. We defined a tract to be a "high lead tract" if at least 1% of the tests in the tract resulted in elevated blood lead levels (meaning 5+ mcg lead/dL). This left us with 106 "high lead" tracts and 585 "safe" tracts. The location of these "high lead" tracts in the Twin Cities metropolitan area can be seen below. It is clear that the majority of them fall in the Minneapolis-Saint Paul city limits. 

```{r}
river_lakes_big <- river_lakes %>%
  filter(AREA_ACRES >= 500)
roads_metro <- st_crop(roads,st_bbox(lead_census))
lead_census %>%
  ggplot()+
  geom_sf(aes(fill=as.factor(HighLead)), lwd = 0.2, color = "white")+
  geom_sf(data = river_lakes_big, fill = "lightblue",color = "lightblue")+
  theme_classic()+
  labs(title = "Elevated blood lead levels in the Twin Cities", subtitle = "<strong><span style='color:red'>Red tracts have at least 1% of children tested with high elevated blood levels</span></strong></b>")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        legend.position = "none", 
        plot.title.position = "plot", 
        plot.title = element_markdown(family = "mono"), 
        plot.subtitle = element_markdown(family = "mono", size = 8))+
  scale_fill_manual(values = c("navy", "red"))
```

The reason why this research question is important is because understanding what is correlated with tracts having high lead levels can help the Minnesota Department of Health, organizations, and families protect children from lead exposure. For example, it wouldn't be unreasonable to expect tracts with older homes to have higher lead levels, as these homes are more likely to have been built when science did not know the harms of lead pipes and paint. On March 28, 2022, Saint Paul Mayor Melvin Carter announced a $14.5 million American Rescue Plan investment to remove thousands of lead pipes across the city [@saintpaul]. If home age appears a strong indicator of high lead levels, identifying tracts with old homes, high lead levels, and lots of young children can alert the city to replace their pipes first. In our research we also might search for a relationship between testing, income, and lead levels. If we are to find certain income groups getting tested more or less than others holding other variables constant, we can shed light on that and advocate for resources to get specific tracts the testing they need and deserve given their exposure. 

To help us understand what is correlated with a tract being "high lead", we will need more than just the information provided by the MDH of tract lead levels. Using the **tidycensus** [@tidycensus] package in R, we can access a plethora of information on each census tract including its estimated mean age, mean income, population, proportion of family households, home age, and so much more. We begin by exploring the relationship between many of these variables and testing as well as EBLLs.

# Exploratory Data Analysis 
