---
title: "Data Visualization"
author: "Erin Franke and Nicholas Di"
date: "4/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidycensus)
library(tidyverse)
library(survey)
library(srvyr)
library(sf)
library(ggtext)

load("DataShapefiles/lead_spatial.RData")
river_lakes <- st_read("DataShapefiles/shp_water_lakes_rivers")
roads <- st_read("DataShapefiles/tl_2019_27_prisecroads")
roads_sub <- st_crop(roads,st_bbox(lead_census))
```

Where our analysis is taking place: 
```{r}
lead_census %>%
  ggplot()+
  geom_sf(fill = "navy", lwd = 0.2, color = "white")+
  geom_sf(data = roads, fill = "lightblue",color = "lightblue")+
  theme_classic()+
  labs(title = "Twin Cities Metropolitan region on MN road map")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        legend.position = "none", 
        plot.title.position = "plot", 
        plot.title = element_markdown(family = "mono"), 
        plot.subtitle = element_markdown(family = "mono", size = 8))+
  scale_fill_manual(values = c("navy", "red"))
```

Start by looking at where lead levels are high:
```{r}
river_lakes_big <- river_lakes %>%
  filter(AREA_ACRES >= 500)
roads_metro <- st_crop(roads,st_bbox(lead_census))
lead_census %>%
  ggplot()+
  geom_sf(aes(fill=as.factor(HighLead)), lwd = 0.2, color = "white")+
  #geom_sf(data = roads_metro, fill = "green",color = "green", lwd= 0.3)+
  geom_sf(data = river_lakes_big, fill = "lightblue",color = "lightblue")+
  theme_classic()+
  labs(title = "Elevated blood lead levels in the Twin Cities", subtitle = "<strong><span style='color:red'>Red tracts have at least 1% of children tested with high elevated blood levels</span></strong></b>")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        legend.position = "none", 
        plot.title.position = "plot", 
        plot.title = element_markdown(family = "mono"), 
        plot.subtitle = element_markdown(family = "mono", size = 8))+
  scale_fill_manual(values = c("navy", "red"))
  #scale_fill_viridis_c(labels = scales::comma)
```

Visualize a few different variables

```{r}
lead_census %>%
  ggplot()+
  geom_sf(aes(fill = medincomeE), lwd = 0.2, color = "white")+
  theme_classic()+
  labs(title = "Median family income by Twin Cities census tract", fill = "")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        plot.title.position = "plot", 
        plot.title = element_markdown(family = "mono"))+
  scale_fill_viridis_b(labels = scales::comma)

lead_census %>%
  ggplot()+
  geom_sf(aes(fill = propFamilyHouseholds), lwd = 0.2, color = "white")+
  theme_classic()+
  labs(title = "Proportion of family households by Twin Cities census tract", fill = "")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        plot.title.position = "plot", 
        plot.title = element_markdown(family = "mono", size = 9))+
  scale_fill_viridis_c()

lead_census %>%
  ggplot()+
  geom_sf(aes(fill = propHomesBuiltPre1950), lwd = 0.2, color = "white")+
  theme_classic()+
  labs(title = "Proportion of homes built before 1950 by Twin Cities census tract", fill = "")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        plot.title.position = "plot", 
        plot.title = element_markdown(family = "mono", size = 9))+
  scale_fill_viridis_c()

lead_census %>%
  ggplot()+
  geom_sf(aes(fill = (propHomesBuiltPre1950+propHomesBuilt1950to1969)), lwd = 0.2, color = "white")+
  theme_classic()+
  labs(title = "Proportion of homes built before 1970 by Twin Cities census tract", fill = "")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        plot.title.position = "plot", 
        plot.title = element_markdown(family = "mono", size = 9))+
  scale_fill_viridis_c()

lead_census %>%
  ggplot()+
  geom_sf(aes(fill = medageE), lwd = 0.2, color = "white")+
  theme_classic()+
  labs(title = "Estimated median age of Twin Cities census tracks", fill = "")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        plot.title.position = "plot", 
        plot.title = element_markdown(family = "mono", size = 9))+
  scale_fill_viridis_c()
```

Looking specifically at some high lead tracts in Ramsey and Hennepin County:
```{r}
lead_ramsey <- lead_census %>%
  filter(primary_county == "Ramsey County") 
roads_ramsey<- st_crop(roads,st_bbox(lead_ramsey))
st_crs(river_lakes) <- st_crs(lead_ramsey)
#water_ramsey <- st_crop(river_lakes, st_bbox(lead_ramsey))
lead_ramsey %>%
  ggplot()+
  geom_sf(aes(fill=percent), lwd = 0.2, color = "white")+
  geom_sf(data = roads_ramsey, fill = "green",color = "green", lwd= 0.3)+
  theme_classic()+
  labs(title = "Proportion of children tested with elevated lead", subtitle = "blood levels in Ramsey County", fill = "")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        #legend.position = "none", 
        plot.title.position = "plot", 
        plot.title = element_markdown(family = "mono", size=8),
        plot.subtitle = element_markdown(family = "mono", size=8),
        legend.title = element_text(family="mono"))+
  scale_fill_viridis_b(option = "inferno")

lead_hennepin <- lead_census %>%
  filter(primary_county == "Hennepin County") 
roads_hennepin<- st_crop(roads,st_bbox(lead_hennepin))
#water_hennepin <- st_crop(river_lakes, st_bbox(lead_hennepin))
lead_hennepin %>%
  ggplot()+
  geom_sf(aes(fill=percent), lwd = 0.2, color = "white")+
  geom_sf(data = roads_hennepin, fill = "green",color = "green", lwd= 0.3)+
  theme_classic()+
  labs(title = "Proportion of children tested with elevated lead", subtitle = "blood levels in Hennepin County", fill = "")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        #legend.position = "none", 
        plot.title.position = "plot", 
        plot.title = element_markdown(family = "mono", size=8),
        plot.subtitle = element_markdown(family = "mono", size=8),
        legend.title = element_text(family="mono"))+
  scale_fill_viridis_b(option = "inferno")

lead_dakota <- lead_census %>%
  filter(primary_county == "Dakota County") 
roads_dakota<- st_crop(roads,st_bbox(lead_dakota))
#water_dakota <- st_crop(river_lakes, st_bbox(lead_dakota))
lead_dakota %>%
  ggplot()+
  geom_sf(aes(fill=percent), lwd = 0.2, color = "white")+
  geom_sf(data = roads_dakota, fill = "green",color = "green", lwd= 0.3)+
  theme_classic()+
  labs(title = "Proportion of children tested with elevated lead", subtitle = "blood levels in Dakota County", fill = "")+
  theme(axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        #legend.position = "none", 
        plot.title.position = "plot", 
        plot.title = element_markdown(family = "mono", size=8),
        plot.subtitle = element_markdown(family = "mono", size=8),
        legend.title = element_text(family="mono"))+
  scale_fill_viridis_b(option = "inferno")
```

Looking at children tested using number of children in the tract aged 0-5. It may be better to use a certain number of tests as a point of whether a tract should be included or not as it seems somewhat common for there to be more tests in a tract than kids of testing age. 
```{r}
lead_census %>%
  mutate(prop_tested = tested/numChildtestingage) %>%
  select(prop_tested, tested, numChildtestingage) %>%
  arrange(desc(prop_tested))
```

Looking at testing variable and relationships:
- High income census tracts tend to not have high lead \
```{r}
lead_testing <- lead_census_clean %>%
  st_drop_geometry() %>%
  mutate(testRatio = tested/numChildtestingage, 
         CensusAgeE = propHomesBuiltPre1950 * runif(1, 72,122) + propHomesBuilt1950to1969*62.5 + propHomesBuilt1970to1989*42.5 + propHomesBuilt1990to2009*22.5 + propHomesBuilt2010tonow*6, highIncome = case_when(medincomeE > 100000 ~ "highInc",
                                TRUE ~ "lowerInc"),
         Homecat = cut(CensusAgeE, 3), 
         multipleTest = case_when(testRatio > 1 ~ "manyTest",
                                  TRUE ~ "lessTest"))
  
lead_testing %>%
  ggplot(aes(x=medincomeE, y=testRatio, color=as.factor(HighLead)))+
  geom_point()

lead_testing %>%
  ggplot(aes(x=CensusAgeE, y=testRatio, color=as.factor(HighLead)))+
  geom_point()

# new homes get tested less often, especially with low lead levels
# older homes get tested more, especially when with higher lead levels
lead_testing %>%
  tabyl(Homecat, multipleTest, HighLead)

# if you have high lead, income isn't necessarily a factor for how much you are tested
lead_testing %>%
  filter(HighLead == 1) %>%
  ggplot(aes(x=medincomeE, fill = multipleTest))+
  geom_density(alpha=0.5)+
  theme_classic()+
  scale_fill_manual(values = c("red", "navy"))

lead_testing %>%
  group_by(HighLead, multipleTest) %>%
  summarize(mean(NumHouseE),mean(CensusAgeE), mean(medincomeE), mean(PopE), mean(medageE))
```
